---
title: "Day 4: Building Websites"
output:
  html_document: 
    toc: yes
    toc_float: yes
    theme: flatly
    css: style.css
---

:::{.blue}
**Learning Objectives**

* Recognize the minimal requirements to build a website using R and Markdown. 
* Build three types of websites using R Studio, Github, Markdown and a sprinkle of Hugo
* Publish these website to GitHub and/or Netlify
:::

----

# Introduction

## Motivating Examples

* Static pages
    - Personal website: http://www.emilyzabor.com/
    - Academic website: http://webpages.csus.edu/wiscons/ 
    - Academic website: http://webpages.csus.edu/fitzgerald/about/
    - Workshop: https://psyteachr.github.io/hack-your-data/index.html 
    - Course materials https://norcalbiostat.github.io/MATH315/index.html
    - Project report https://norcalbiostat.github.io/chem_ss/
* Static pages + blog
    - Course website for Data Analytics for Social Good https://norcalbiostat.github.io/DATA185/ 
    - Piping hot data https://www.pipinghotdata.com/ 
    - Teaching in Production https://rstudio-education.github.io/teaching-in-production/ 
* Blog centric
    - https://alison.rbind.io/
    - http://www.citizen-statistician.org/
    - https://beyondrawdata.rbind.io/
    - https://chicodatafest.netlify.com/
    - http://datascience.csuchico.edu/
    - https://yihui.org/
    - https://robjhyndman.com/ 
    - https://www.norcalbiostat.com/ 
    

## Prerequisites
To full participate in today's activites, you must have the following: 

* **Version Control:** A github account, git installed on your computer, the ability to push and pull from a local repository to a remote one, and the ability to connect git repositories to R through projects. 
* **R & R Studio:** with the following packages: `knitr`, `blogdown v1.3`, `distill v1.2.4`, 
    - use blogdown to install Hugo by typing `blogdown::install_hugo()`
* **A [Netlify](https://www.netlify.com/) account.** I recommend that you sign up with your github account.  
* **Set Intentions** It's easy to get lost in all the settings, tweaks, themes, customizations and so forth. Decide on some clear objectives to help keep you on track.  


:::{.green}
**Exercise 1: Set your intentions (Think-Pair-Share)**

1. Write down three things you want to accomplish with a website
2. Get into random breakout rooms with another persons
3. Share & discuss your goals
:::

<br>

:::{.blue}
**First watch, then do**

We will be alternating a lot today where you watch me demo first, then you replicate on your own or in groups.
:::

----

# Getting started: RMarkdown

R Markdown websites like this one can be built with a minimum of two files: a `_site.yml` and an `index.Rmd`. 

The `_site.yml` file controls the overall structure of the website, such as the order of the navigation bar and the color theme. 

When these files are _rendered_ into HTML files, they can be hosted on GitHub for free. Then after turning on a few settings in your GitHub repository (repo) these files can be viewed as a magnificent webpage. 




### 1. Create a public Github repo. 
We are going to employ the "[GitHub first, R Studio second](https://happygitwithr.com/new-github-first.html)" approach. 

* From your *Repository* page, click the green <span style="color:green">[NEW]</span> button.  
* (_Suggestion_): Name this repo something like `website_test_rmarkdown` so you know it's a playground.
* keep it a PUBLIC repo
* YES initialize with a README

Click 'Create Repository' to finish, but keep this page open. We'll come back to it a few times.

### 2. Create a project in R Studio and clone your new repo into this project. 

* Top right corner in R Studio click: *New Project --> Checkout from version control --> Git*
* Go back to GitHub, and click on the green <span style="color:green">[Clone or Download]</span> button.
  - Click the clipboard icon to copy the URL to your repo. 
  - Paste this URL into the _Repository URL_ box in R Studio. 
    * **(_Strongly Suggested_)**: Name this project the same as the repo name.
    * Subdirectory: Your path, your choice. Just don't put a repo inside a repo. 
* Click _Create Project_

### 3. Create necessary files in R Studio.

* In the *Terminal* window (tab to the right of the *Console*), create the `_site.yml` and `index.Rmd` files using the `touch` command. Type the following commands in the Terminal one at a time, pressing <kbd>Return</kbd> or <kbd>Enter</kbd> after each line to execute. 
```
touch _site.yml
touch index.rmd
```
You can view these files in the _File_ pane (lower right). You can open supported file types by clicking on the name of the file from this pane. 

* Open both the `_site.yml` and `index.Rmd` files now. 

:::{.green}
`r emo::ji("group")` Do steps 1-3 on your own. 
Put up a `r emo::ji("heavy_check_mark")`when you are done. 
:::

### 4. Configuring the `_site.yml` file. 
This is the file that controls your overall site's configuration. If you are familiar with R Markdown files, this is the same type of 'language' that is used in the header area to control the document structure. 

YAML headers, and this file, are very particular about spacing and tabs. To avoid unnecessary grief we are going to copy/paste code from the collaborative notes. 

```
name: "my-website"
output_dir: "."
navbar:
  title: "My Website"
  left:
    - text: "Home"
      href: index.html
```

* Give your website a name and title!


### 5. Configure your index page. 

This is your landing page. That is, the first page people see when they go to your website. You can add a title to this page using a YAML header as is shown in the bookdown website, or you can use a pound sign `#` as a first level header to make this title. 

* Create a title for your page using *either* a YAML header or a `#` sign. Using both is redundant. 
* Write a sentence or two welcoming your viewers to your website. 


### 6. Building & previewing your website locally. 

Since we created the `.yml` file manually after creating the R Project, your project doesn't quite know yet that it has a website to build. 

* Save your work and close down R studio. 
* Navigate to your repository folder on your computer, and open the project by double clicking on the R project file icon (a cube with the letter R). 
    - _Note: This is always how you should open your R Project files_

There should now be a **BUILD** tab in the top right pane.

* Click this tab, and then click the _Build Website_ button. 

An HTML file should now appear in your viewer pane. 

* Click the 'Show in new window' button to see your site in a full browser window. 

* Make a change to your `index.Rmd`, knit and refresh your browser window.

:::{.green}
`r emo::ji("group")` Do steps 4-6 on your own. 
Put up a `r emo::ji("heavy_check_mark")`when you are done. 
:::

<br>

:::{blue}
**Notes, Tips and Comments** 

* `r emo::ji("warning")` <span style="color:red">Clicking "Build Website" will render every `.md` and `.Rmd` file in your top level project folder into an HTML file. </span>
    - This is important to keep in mind when creating a website with a lot of pages. 
    - Code files in sub-folders will not be automatically rendered. 
* You can re-render individual files by clicking 'Knit'. 
* You can leave this browser window open as you work, refreshing the page to see new changes. 
:::

### 7. Push to GitHub & publish your website. 

Let's get your test site up for the world to see. 

* In the _Terminal_ window, stage and commit your files to version control by executing the following commands:

```
git add <file name>
```

This will _stage_ all files that have been changed and/or added. _Staging_ is the process of adding a file to be tracked under version control. 

```
git commit -m "first commit"
```

Every commit needs a message. Try to make it informative, yet brief. 

```
git push
```

This will _push_ your changes up to GitHub's servers online. You may have to enter your github username and password here. See [happy git with R](https://happygitwithr.com/push-pull-github.html) for help storing your github credentials. 

* Go back to your GitHub repo page and refresh - you should now see that your repo now contains the new code, and HTML files that you have been working on. 
* Click _Settings_ in the top right of the navigation bar in your repository. 
    - Scroll down to the **Pages** section in the left navigation pane. 
    - Under _Source_, select _main_
    
After the page refreshes, you will see the URL to your website in at the top of GitHub pages section. 

* Click on this link to see what your public facing website looks like!
* Copy this URL, we're going to paste it in the repo settings so it can be found easy. 
* Go back to your main repo page by clicking on the name of your repo at the top of the GitHub page. 
    - Click **About** on the right side panel
    - Paste the URL into the _Website_ field. 

### 8. Adding Content. 

Let's add some content.  

* Create a new R Markdown file using _File -> New R Markdown_ so we can get the example R Markdown content that includes a plot. 
    - Save this file with a name such as `project.Rmd`
* Add a link to the project HTML page to your `_site.yml` configuration file by adding a new `text` line to the `navbar`.   

```
    - text: "My Projects"
      href: project.html
```

`r emo::ji("warning")` <span style="color:red">Be mindful about the spacing. </span> Make sure the `- text:` is directly in line with the home page item in the YAML header. 


* Rebuild your website by clicking the _Build Website_ button in the _Build_ tab in the top right pane.
* View your changes locally, make a change if you see fit. 
* Add, commit, and push your changes. 
* Refresh your live webpage on GitHub to see the results! 


:::{.green}
`r emo::ji("group")` Do steps 7-8 in your breakout rooms. 
Come back to the main room when you are done. 
:::

### Additional Features & References

* Drill down navigation bar - indent in the YAML header under `menu`
* Right alignment of some text
~~~md
---
name: "my-website"
output_dir: "."
navbar:
  title: "My Website"
  left:
    - text: "Home"
      href: index.html
    - text: "Courses"
      menu:
      - text: "Math 1"
        href:  math1.html
      - text: "Math 2"
        href: math2.html
  right: 
    - text: "About"
      href: about.html
---
~~~

* R Markdown sites from R Studio https://rmarkdown.rstudio.com/lesson-13.html
* Chapter 10.5 in the Blogdown book https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html 

----

# Level up: Blog Aware with Distill
Distill based websites are a nice enhancement to the Rmarkdown website. Very similar setup/structure, but with added ability to manage a blog. You can actually create static Websites with distill as well. [Learn more about the difference here](https://rstudio.github.io/distill/blog.html#website-or-blog).

### 1. Create a new Github Repo and connect it to an R Project. 

Follow steps 1 & 2 from the section above. 

### 2. Use `distill` to turn the project into a website. 
In the console, type the following: 

```{r, eval=FALSE}
library(distill)
create_blog(dir=".", title = "My Blog")
```

The basic scaffolding for a blog and an initial welcome post will be created within the root (`.`) sub-directory.

* Shut down RStudio and re-open the project to enable the _build_ tab. This is similar to step 6 above.  

### 3. Configure the YAML header and build your site.  
Open `_site.yml` Looks similar eh? 

* output: `distill::distill_article` instead of `html_document`
* Change the description and name. 
* Change the `output_dir:` in the YAML to  `"docs"`
* Build your website by using the _build_ tab and see if you can spot where these items show up. 
    - You can also run `render_site()` in the console to build your site instead of using the _build_ tab. 

:::{.red}
#### Changing the output directory

* For this example we will be publishing through github pages, which doesn't seem to recognize `_site`, but does `docs`.  
* This seems to be a github thing. Not a universal thing
:::

What's happening, is that all `.md` and `.Rmd` files in your root directory get knitted into `.html` files and placed into the `docs` folder. All content in the `_posts` folder also gets **directly copied** into the `docs` folder.

### 4. Create a post
Run the following code in the console: 

```{r, eval=FALSE}
create_post("Hello World", draft=TRUE)
```

* Posts are created in the `_posts` subdirectory. 
* An entire subdirectory is created that reflects the creation date and name of post: `2021-06-15-hello-world`
* You can work on these posts individually, knitting as needed. 
    - Files in a sub-directory do not get re-rendered when you build your site. 
    - _See [this section](https://rstudio.github.io/distill/blog.html#renaming-posts) of the distill documentation for how to rename posts and directories_

* When you are ready to make your post live, either fully remove or update the `draft: true` section in the YAML header


:::{.green}
`r emo::ji("group")` In breakout rooms: 

1. Create new static page
2. Add the new static page to the navigation bar
3. Create a new post and write a few sentences about what you want to use this site for. 
4. Build your site and make sure you see all changes. 

Come back to the main room when you are done. 
:::

### 5. Publishing your website
* Add a `.nojekyll` file using either `touch .nojekyll` in the terminal, or `file.create(".nojekyll")` in the R console
    - Github pages tends to want to use the Jekyll website generator to compile pieces of a website into one cohesive site. You don't need that in this case. 
* Add, commit, and push your changes to your github repository.
* Activate github pages, and publish from the from the `main` branch and the `/docs` folder. 


:::{.green}
#### `r emo::ji("group")` RANDOM SHARE OUT: 

* You will be randomly assigned a new breakout room.  
* Each person take 1 minute, share their screen and show others what their page looks like. 
:::


### Additional References
This was a basic introduction to these types of websites. See the documentation at https://rstudio.github.io/distill/blog.html to see how to: 

* create custom listings for subsets of posts
* add categories to posts so they can be searched through
* allow Disqus commenting and sharing of posts on social media
* Themes
* Publishing on sites other than github like Netlify
* and more


----

# Full Fancieness: Hugo `r emo::ji("warning")` Section in development

The "simple" website that was built above can be extended and enhanced well enough with some CSS and Rmarkdown wizardry. However, sometimes you may want a little more of a 'modern', 'dynamic' or 'responsive' feel, or to include a blog in your website. Here are some examples:  

* https://georgecushen.com/


This is when we enter the realm of [[Hugo]](https://gohugo.io/), - the self proclaimed "most popular open-source static site generator". 

Hugo is it's own language, and for advanced usage of Hugo-based websites, some understanding of how Hugo works and reading _those_ docs may be required. 

For now, we are just going to show you how to get started in this realm of fancy-ness using the R package `blogdown`. Then you can go break it later. The super official blogdown documentation is linked in the References section at the bottom of this page.  

If you did not do so beforehand, install the `blogdown` package, and use it to install Hugo using `blogdown::install_hugo()`. 

### 1. Create a new Github Repo and connect it to an R Project. 

You got this.  

### 2. Create a new site with a specified theme

The `blogdown` package will setup the necessary files, folder structures and theme files from a pre-specified them that is available on github. The example we will start with is a very simple site containing a few pages and a blog. 


```{r, eval=FALSE}
blogdown::new_site(theme='yihui/hugo-lithium')
```

This could take up to a minute to download all necessary files. 

When it is done, your website will be rendered and appear in the bottom right viewer pane. Click the 'Show in new window' to open it in your full browser. 


### 3. Explore configuration files. 
Let's start by looking at the files that are contained in the root folder of this project. 

* `index.Rmd`: Nothing really is there! This is one of the primary differences between this type of site building and the one we just finished. Content is handled in a very different manner. Hugo + Blogdown aim to help you focus on content creation and these programs put the pieces together for you.

* `config.toml`: Functions similar to the `_site.yml` file, you can think of TOML as having the same functionality as the YAML, but with different syntax. TOML uses a key-value pair of options, such as

```
title = "A Hugo website"
```

Areas of the website, like the menu (navbar), are controlled in sections that start with brackets and look like the following: 

```
[[menu.main]]
    name = "About"
    url = "/about/"
```

**Do together** 

* Remove the twitter link
* Change the GitHub link to point to your GitHub account
* Give your site a different title. 

`r emo::ji("question")` Were you able to see the results of each of the changes you made on the live site? 


### 4. Explore and modify static content files. 

Open the `content` folder. Right now there is only an `about.md` file. **Open this file**. 

We see a familiar YAML header, and a body of text. 

**Do together** 

* Change the title and content. 
* Save this file, go back to your local site and refresh the live preview to see the changes. 

This folder is where we can add more static pages, such as a CV. Since we are not using any R code here, we can stick with the simpler markdown only file format. 

**Do together** 

* Copy the `about.md` file, rename this as `cv.md`. 
  - You can do this in R Studio by clicking the box to the left of the file name in the Files window, then clicking _More --> Copy_
* Change the title, add a few bullet points and save. 
* Go back to the `config.toml` file, and add a new ``[[menu.main]]` section as shown below so that this link shows up in the navbar.

```
[[menu.main]]
    name = "CV"
    url = "/cv/"
```


### 5. Blog posts

On your newly updated website, clicking on the icon in the top left. This will take you back to the landing page. For this theme, the landing page is a blog (instead of a static page such as `About`). Clicking on one of these posts takes you to the page that contains the post itself. 

The content files for posts are stored under `content/post/` folder. If you look there now you'll see some markdown (`.md`), some R Markdown (`.Rmd`) files, and their associated `.html` output files. 

**Creating new blog posts**

* Option 1: Copy/paste/modify one of the example files already in this folder. 
* Option 2: Use the R Studio Addins to setup the page structure for you. 
    - **This is the one we're going to use**
    - Using the UI allows you to easily set catgories, tags and the like. 
    
**Do together** 
    
* Click _Addins --> New Post_
  - This will likely require you to install and/or update new packages including (but not limited to) 
    `fastmap`, `miniUI`, `htmltools`, `shiny`. 
  - If you run into seemingly endless package installation loops, restart RStudio and re-try. 
* Give your post a title, author name, a category and a tag. 
    - Categories and tags are optional, but for this example go ahead and set them. 
* Select the `.Rmd` format for this example. 
* Select <kbd>Done</kbd>

Let's do a quick analysis of the `cars` data set to add some simple content for display. 

* Don't knit this file. Just save. 

_(Instructors note: Intro, summary, pander, sentence, plot, sentence, inline R, hide code)_


### 6. Serve Site

We've been "cheating" a little so far in that our site has already been "Served". 
When we started the new site in step 2, not only did `blogdown` download the theme files and exmaple site, it also ran `blogdown::serve_site()`. This is what allows for the live preview. 

At a later point when you want to come back to working on your website, opening the project file itself will not initiate a `serve_site()`. You will have to do this manually. Again, there is an **Addin** for this. 

### 7. Push to GitHub. 

Go ahead and save your work, add, commit, and push your content to your github account. 


### 8. Hosting / Deploying your website. 

If you have your own server, and you are familiar with website deployment then you can roll your own as you are used to. For those of us that have absolutely no idea how to host a website, a very good option that is very user friendly (and free) is [Netlify](https://www.netlify.com/). 

You *could* use GitHub Pages as we did previously to host a Hugo based website, but there are some tweaks that you have to do to get GitHub to work with/around Jekyll in the way Blogdown/Hugo builds the sites. If you want to learn more about the why's, go read the docs at https://bookdown.org/yihui/blogdown/deployment.html

**Do together (Time pending)**



### Closing comments

* Hugo has a TON of themes: https://themes.gohugo.io/ . Not all work with `blogdown`, and not all are easy to work with. Here is a blog post by Peter Baumgartner on [important thoughts on choosing a theme to work with](https://portfolio.peter-baumgartner.net/2018/07/19/blogdown-using-themes/). 
* The theme shown here has a very simple TOML configuration file. Each theme is different and unique. Reading the documentation (generally contained in the readme) is critical. 


### Additional References


* `blogdown`: Creating Websites with R Markdown https://bookdown.org/yihui/blogdown/ , https://github.com/rstudio/blogdown 
    * Getting started with the Academic theme: https://sourcethemes.com/academic/docs/get-started/
    * Up and running with blogdown in 2021 https://alison.rbind.io/blog/2020-12-new-year-new-blogdown/ 

    
